---
title: Homework 1
subtitle: "`ggplot2` & `dplyr` Fundamentals  · Quarto Blogging"
author: Makene_Comeau
date: last-modified
from: markdown+emoji
execute: 
  eval: false
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 5
---

```{r}
#| include: false
library(tidyverse)
library(skimr)
```


# 📌 Directions

- Submit **one Quarto document** (`.qmd`) to Brightspace:

  - `danl-310-hw1-LASTNAME-FIRSTNAME.qmd`  
    (e.g., `danl-310-hw1-choe-byeonghak.qmd`)

- **Due:** **February 16, 2026, 2:00 P.M. (ET)**

- For **visualization questions**, you must provide:
  1) the **`ggplot2` code**, and  
  2) a **written comment** (2–4 sentences) interpreting the corresponding figure.

- Unless a question says otherwise, use `dplyr` verbs (`filter()`, `select()`, `mutate()`, `group_by()`, `summarise()`, `arrange()`, `count()`, etc.) and `ggplot2`.

<br>

# ✅ Setup

```{r}
#| echo: true
library(tidyverse)
library(skimr)
```

---

<br><br>

# Part 1. Data Visualization & Summaries

## A. Orange Juice Promotions (`oj`)

Use the following dataset for Questions 1–6.

```{r}
#| echo: true
oj <- read_csv("https://bcdanl.github.io/data/dominick_oj_na.csv")
```

### Question 1. Quick inspection
1. Print the first 10 rows of `oj`.
2. Use `skimr::skim()` (or another method) to summarize the variables.

```{r}
library(skimr)
skim(oj)
oj %>%
  head(n=10)
```

<br>

### Question 2. Brand-level descriptive statistics (filter + group summary)
Filter to the three brands: **tropicana**, **minute.maid**, and **dominicks**. Then compute **mean** and **standard deviation** of:

- `sales`
- `price`

for each brand.

```{r}
filter_brands <- oj %>%
  filter(brand %in% c("tropicana", "minute.maid", "dominicks")) %>%
  group_by(brand) %>%
  summarise(
    mean_sales = mean(sales, na.rm = TRUE),
    sd_sales = sd(sales, na.rm = TRUE),
    mean_price = mean(price, na.rm = TRUE),
    sd_price = sd(price, na.rm = TRUE)
  )

print(filter_brands)

```

Write 2–3 sentences comparing the three brands based on your results.

<br>

### Question 3. Remove missing values
Create a new data frame `oj_no_na` that removes observations with **missing** values in **either** `price` or `sales`.

1. Show the number of rows in `oj` and `oj_no_na`.
2. Report how many rows were removed.

```{r}
oj_no_na <- oj %>% drop_na(price,sales)

nrow(oj)
nrow(oj_no_na)
# 28947
# 28941

nrow(oj) - nrow(oj_no_na)
# 6 row removed 

```

<br>

### Question 4. Price distribution by brand (ggplot + comment)
Using `oj_no_na`, make a figure that compares the **distribution of `price` across `brand`**.

- Choose ONE main approach (e.g., boxplot, violin plot, density plot, histogram with facets).
- Include clear axis labels and a readable title.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(oj_no_na, aes(x = brand, y = price, fill = brand)) +
  geom_boxplot(width = 0.2, fill = "white", alpha = 0.7) + 
  labs(
    title = "Distribution of Price Across Different Brands",
    x = "Brand",
    y = "Price"
  ) 
 
```

**(b) Comment (2–3 sentences):** What differences do you see across brands (center, spread, skewness, outliers, etc.)?
# Tropicana has the highest price of all brands followed by minute maid and dominicks. Dominicks has high and low price outliers. 

<br>

### Question 5. Log–log price–sales relationship by brand (ggplot + comment)
Using `oj_no_na`, visualize how the relationship between:

- `log10(sales)` and `log10(price)`

varies by **brand**.

- Use a scatter plot with transparency (e.g., `alpha = 0.3`).
- Add a fitted line (e.g., `geom_smooth(method = "lm", se = FALSE)`), and
- Use faceting OR color to distinguish brands.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(data = oj_no_na, aes(x = log10(price), y = log10(sales))) +
  geom_point(alpha = 0.2) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  facet_wrap(~ brand) + 
  labs(title = "Log-Log Price-Sales Relationship by Brand",
       x = "log10(Price)",
       y = "log10(Sales)")  
```

**(b) Comment (2–3 sentences):** Do you see evidence that higher prices are associated with lower sales? Does the pattern differ by brand?

# Lower prices does associate with high sales. The higher the price the less the sales are in each brand it is clear. It is most clear in the brand dominicks. 

<br>

### Question 6. Log–log relationship by brand and ad status (ggplot + comment)
Now extend Question 5 by incorporating `ad_status`. Visualize how the relationship between:

- `log10(sales)` and `log10(price)`

varies by **brand** *and* **`ad_status`**.

- Use faceting (`facet_grid()` or `facet_wrap()`), and/or
- Use color/shape for `ad_status`.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(data = oj_no_na, aes(x = log10(price), y = log10(sales), color = ad_status)) +
  geom_point(alpha = 0.5) + 
  labs(title = "Log10(Sales) vs Log10(Price) by Ad Status",
       x = "Log10(Price)",
       y = "Log10(Sales)",
       color = "Ad Status") +
  facet_wrap(~ brand) 
```

**(b) Comment (2–3 sentences):** How does advertising status appear to shift the relationship (level shift, slope change, or no clear difference)?

# Adding the ad status does not seem to change the shift of sales and price.It does show that with ads there are more sales with still a lower price. 
---

## B. MLB Batting Trends (`mlb_bat`)

Use the following dataset for Questions 7–8.

```{r}
#| echo: true
mlb_bat <- read_csv("https://bcdanl.github.io/data/MLB_batting.csv")
```

### Question 7. Create yearly hit percentages (data transformation)
Create a data frame `mlb_hit_pct` that contains **yearly hit percentages** for each `hit_type`
(**Single**, **Double**, **Triple**, **HomeRun**).

Your final table should have:

- `year`
- `hit_type`
- `hit_pct` (a percentage or proportion)


```{r}
library(dplyr)

mlb_hit_pct <- mlb_bat %>%
  filter(hit_type %in% c("Single", "Double", "Triple", "HomeRun")) %>%
  count(year, hit_type, percentage, name = "hits") %>%
  group_by(year) %>%
  ungroup() %>%
  select(year, hit_type, percentage)

mlb_hit_pct
```

<br>

### Question 8. Visualize trends (ggplot + comment)
Make a figure that shows the yearly trends in hit percentages for each `hit_type`.

**(a) Provide your `ggplot2` code:**

```{r}
library(ggplot2)


ggplot(mlb_hit_pct, aes(x = year, y = percentage, color = hit_type)) +
  geom_line(size = 1) +
  geom_point(size = 0.2) +
  labs(
    title = "Yearly Trends in MLB Hit Percentages by Hit Type",
    x = "Year",
    y = "Hit Percentage",
    color = "Hit Type")
```

**(b) Comment (2–3 sentences):** Which hit types are increasing/decreasing over time? Any notable breaks or eras?

# Single hit has the highest hit percentage. Double hit increases slowly overtime while the triple hit decreases and seems rare. 

## C. Titanic Survival (`titanic`)

Use the following dataset for Questions 9–13.

```{r}
#| echo: true
titanic <- read_csv("https://bcdanl.github.io/data/titanic_cleaned.csv")
```

### Question 9. Two-way count table
Create `titanic_class_survival` that counts passengers by `class` and `survived`.

```{r}
titanic_class_survival <- titanic %>%
  count(class, survived)
titanic_class_survival
  
```

<br>

### Question 10. Age distribution by class and gender (ggplot + comment)
Visualize how the **distribution of `age`** varies across `class` and `gender`.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(titanic, aes(x=age, fill=gender))+
  geom_density(alpha = 0.5, na.rm = TRUE)+
  facet_wrap(~class)+
  labs(
    title = "Age Distribution by Class and Gender",
    x = "Age",
    y = "Density",
    fill = "Gender"
  )
```

**(b) Comment (2–3 sentences):** What differences do you see across class and gender?
# Across class and gender it looks liek there are more males then femals. In the 3rd class it seems the highest age of people is between 20-30.
<br>

### Question 11. Survival rate by class and gender (data + ggplot + comment)
1. Create a summary table with the **survival rate** (proportion survived) by `class` and `gender`.
2. Visualize the survival rates.

**(a) Provide your `dplyr` code for the summary table:**

```{r}
library(dplyr)

titanic_survival_rate <- titanic %>%
  group_by(class, gender) %>%
  summarise(
    survival_rate = mean(survived, na.rm = TRUE),
    .groups = "drop"
  )

titanic_survival_rate
```

**(b) Provide your `ggplot2` code:**

```{r}
ggplot(titanic_survival_rate, aes(x = class, y = survival_rate, fill = gender)) +
  geom_col() +
  labs(
    title = "Survival Rate by Passenger Class and Gender",
    x = "Class",
    y = "Survival Rate",
    fill = "Gender"
  ) 

```

**(c) Comment (2–3 sentences):** Which groups have the highest/lowest survival rates?
# Women in the first class had the highest survival rate. Men had a low survival rate in all classes. 
<br>

### Question 12. Conditional distribution of survived (ggplot + comment)
Create a plot that shows the **distribution of `survived`** across `class` and `gender` using proportions (not raw counts).

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(titanic, aes(x = class, fill = factor(survived))) +
  geom_bar(position = "fill") +
  facet_wrap(~ gender) +
  labs(
    title = "Distribution of Survival by Class and Gender",
    x = "Passenger Class",
    y = "Proportion",
    fill = "Survived"
  ) 
```

**(b) Comment (2–3 sentences):** How does using proportions (instead of counts) change your interpretation?
# Using proportions shows the likelyhood of survival so we can see the gender and class comparison even if the sizes are different. 
<br>

### Question 13. Short interpretation
Write **3 bullet-point insights** supported by your tables/figures in Questions 9–12.
# - Women in the first class had the highest survival rate. Women in the second class also had a high survival rate. 
# - The 3rd class had the youngest ages of people on board. 
# - Men overall had a low survival rate for all ages. 
---

## D. NYC Dog Licenses (`nyc_dogs`)

Use the following dataset for Questions 14–16.

```{r}
#| echo: true
nyc_dogs <- read_csv("https://bcdanl.github.io/data/nyc_dogs_cleaned.csv")
```

### Question 14. Breed frequency table
Create `nyc_dogs_breeds` with:

- non-missing `breed`,
- `n >= 2000`, and
- sorted by `n` descending.

```{r}
nyc_dog_breeds <- nyc_dogs %>%
  filter(!is.na(breed)) %>%
  count(breed, name = "n")%>%
  filter(n>=2000) %>%
  arrange(desc(n))
nyc_dog_breeds
```

<br>

### Question 15. Visualize the distribution of breeds (ggplot + comment)
Using `nyc_dogs_breeds`, make a figure that shows the distribution of breed counts.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(nyc_dog_breeds, aes(x = reorder(breed, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Distribution of Breeds",
    x = "Breed",
    y = "Number of Dogs"
  ) 
  
```

**(b) Comment (2–3 sentences):** Is the distribution concentrated in a few breeds or spread out?
# It is spread out. There are a lot of dog breeds in NYC. 
<br>

### Question 16. “Top breeds” focus (ggplot + comment)
Create a plot of the **top 10 breeds** (by count) and comment on what you observe.

**(a) Provide your `dplyr` + `ggplot2` code:**

```{r}
top10_breeds <- nyc_dog_breeds %>%
  slice_max(order_by = n, n = 10)

ggplot(top10_breeds, aes(x = reorder(breed, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Most Common NYC Dog Breeds",
    x = "Breed",
    y = "Number of Dogs"
  )
```

**(b) Comment (2–3 sentences):** Any surprises? What might explain the pattern?
# It looks like smaller breeds are more common in NYC. That may be because of the condensed area and smaller apartments. 
---

<br><br>

# Part 2. Data Transformation (`nyc_payroll_2025`)

For Questions 17–27, use `nyc_payroll_2025`.

For variable descriptions, see: Citywide Payroll Data (Fiscal Year) on NYC Open Data.

```{r}
install.packages("readr")   # only once
library(readr)

nyc_payroll_2025 <- read_csv("nyc_payroll_2025.csv")
```




### Question 17. Base salary by borough (filter + summarise)
Compute the **mean** and **standard deviation** of `Base_Salary` for workers whose `Work_Location_Borough` is:

- `"MANHATTAN"`
- `"QUEENS"`

Report the two means and two SDs in your write-up.

```{r}
library(dplyr)

salary_borough_summary <- nyc_payroll_2025 %>%
  filter(Work_Location_Borough %in% c("MANHATTAN", "QUEENS")) %>%
  group_by(Work_Location_Borough) %>%
  summarise(
    mean_base_salary = mean(Base_Salary, na.rm = TRUE),
    sd_base_salary   = sd(Base_Salary, na.rm = TRUE),
    .groups = "drop"
  )

salary_borough_summary
```

<br>

### Question 18. High base salary filter
Filter the data to keep only records where `Base_Salary >= 100000`. Report how many rows remain.

```{r}
sal_above_100000 <- nyc_payroll_2025 %>%
  filter(Base_Salary >= 100000)
nrow(sal_above_100000)

# 141464 rows were reported
```

<br>

### Question 19. Distinct agency–title pairs
Select only **distinct** combinations of `Agency_Name` and `Title_Description`.

```{r}
agency_pairs <- nyc_payroll_2025 %>%
  distinct(Agency_Name, Title_Description)
agency_pairs 
```

<br>

### Question 20. Top paid by regular gross pay
Arrange employees by `Regular_Gross_Paid` in **descending** order and show the **top 10** rows with:

- `First_Name`, `Last_Name`
- `Agency_Name`
- `Title_Description`
- `Regular_Gross_Paid`

```{r}
top_10_paid <- nyc_payroll_2025 %>%
  arrange(desc(Regular_Gross_Paid)) %>%
  select(
    First_Name,
    Last_Name,
    Agency_Name,
    Title_Description,
    Regular_Gross_Paid
  ) %>%
  slice_head(n=10)
top_10_paid
```

<br>

### Question 21. Select + rename
Select `Title_Description` and rename it to `Title`. Also select `Agency_Name`, `First_Name`, `Last_Name`, and `Base_Salary`.

```{r}
rename_title <- nyc_payroll_2025 %>%
  select(
    Title = Title_Description,
    Agency_Name,
    First_Name,
    Last_Name,
    Base_Salary
  )
rename_title
```

<br>


### Question 22. Create new pay variables (`mutate()`)
Use `mutate()` to create two new variables:

- `Total_Pay = Regular_Gross_Paid + Total_OT_Paid + Total_Other_Pay`
- `OT_Share = Total_OT_Paid / Total_Pay`

Then show the first 10 rows of:

- `First_Name`, `Last_Name`
- `Agency_Name`
- `Base_Salary`
- `Total_Pay`, `OT_Share`

```{r}
nyc_payroll_2025_1<- nyc_payroll_2025 %>%
  mutate(
    Total_Pay = Regular_Gross_Paid + Total_OT_Paid + Total_Other_Pay,
    OT_Share = Total_OT_Paid / Total_Pay
  ) %>%
  select(
    First_Name,
    Last_Name,
    Agency_Name,
    Base_Salary,
    Total_Pay,
    Work_Location_Borough,
    OT_Share
  ) %>%
  slice_head(n=10)
 nyc_payroll_2025_1 
```

<br>

### Question 23. Borough pay summary (`group_by()` + `summarise()`)
Using the variables you created in **Question 22**, group the data by `Work_Location_Borough` and compute:

- number of employees `n`
- mean `Base_Salary`
- mean `Total_Pay`
- mean `OT_Share` (ignore missing values)

Arrange the results by mean `Total_Pay` in **descending** order and show the summary table.

```{r}
borough_pay_summary <- nyc_payroll_2025 %>%
  mutate(
    Total_Pay = Regular_Gross_Paid + Total_OT_Paid + Total_Other_Pay,
    OT_Share  = Total_OT_Paid / Total_Pay
  ) %>%
  group_by(Work_Location_Borough) %>%
  summarise(
    n = n(),
    mean_base_salary = mean(Base_Salary, na.rm = TRUE),
    mean_total_pay   = mean(Total_Pay, na.rm = TRUE),
    mean_ot_share    = mean(OT_Share, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_total_pay))

borough_pay_summary
```

<br>

### Question 24. Police Department overtime
Filter to `Agency_Name == "POLICE DEPARTMENT"` and arrange by `Total_OT_Paid`.

1. Show the **10 smallest** overtime values.
2. Show the **10 largest** overtime values.

```{r}
nyc_payroll_2025 %>%
  filter(Agency_Name == "POLICE DEPARTMENT") %>%
  arrange(Total_OT_Paid) %>%
  select(First_Name, Last_Name, Title_Description, Total_OT_Paid) %>%
  slice_head(n=10)
```

<br>

### Question 25. Per annum employees
Filter to `Pay_Basis == "per Annum"` and select only:

- `First_Name`, `Last_Name`, `Base_Salary`

```{r}
per_annum_employ <- nyc_payroll_2025 %>%
  filter(Pay_Basis == "per Annum") %>%
  select(First_Name, Last_Name, Base_Salary)
per_annum_employ
```

<br>

### Question 26. Borough then salary
Arrange by `Work_Location_Borough` (ascending) and then `Base_Salary` (descending). Then show the first 15 rows.

```{r}
nyc_payroll_2025 %>%
  arrange(Work_Location_Borough, desc(Base_Salary)) %>%
  slice_head(n=15)
```

<br>

### Question 27. Remove missing last names + count
Remove observations where `Last_Name` is missing (`NA`). Then report the remaining number of rows.

```{r}
nyc_payroll_2025 %>%
  filter(!is.na(Last_Name)) %>%
  nrow()

# 549830 of remaining rows 
```


---

<br><br>

# Part 3. Quarto Blogging (`ice_cream`)

Include the URL to your published blog post.

**Answer:**
# malenecomeau.github.io/


