---
title: "DANL 310 Classwork 6"
editor: visual
---

```{r}
install.packages("nycflights13")
library(ggthemes)
library(tidyverse)

flights <- nycflights13::flights
```

# Question 1
Find all flights that

Had an arrival delay of two or more hours
Flew to Houston (IAH or HOU)
Were operated by United (UA), American (AA), or Delta (DL)
Departed in summer (July, August, and September)
Arrived more than two hours late, but didn’t leave late
Were delayed by at least an hour for the depature, but reduced the delay over 30 minutes during flight
Departed between midnight and 6am (inclusive)

```{r}
not_cancelled <- flights |> 
  filter(!is.na(dep_delay), !is.na(arr_delay))

```

```{r}
# delay more than 2 or more hours 
flights |> 
  filter(arr_delay > 120)

#Flew to Houston (IAH or HOU)
flights |>
  filter(dest %in% c("IAH", "HOU"))

# flights operated by UA, AA, DL
flights |> 
  filter(carrier %in% c("UA", "AA", "DL"))

# departed in summer months jul, aug, sep
flights |>
  filter(month %in% c("7", "8", "9"))

#Arrived more than two hours late, but didn’t leave late
flights |>
  filter(arr_delay > 120,
         dep_delay <= 0)

# Were delayed by at least an hour for the depature, but reduced the delay over 30 minutes during flight
flights |>
  filter(dep_delay >= 60,
         dep_delay - arr_delay > 30) |>
  select(dep_delay, arr_delay)

# Departed between midnight and 6am (inclusive)
flights |>
  fliter((dep_time >= 0 & dep_time <= 600) | dep_time == 2400) |>
  arrange(dep_time)
```

# Question 2

```{r}
# How could you use arrange() to sort all missing values to the start?

flights |>
  skimr::skim(dep_delay)

```

```{r}
flights |>
  arrange(desc(is.na(dep_delay)))
```

# Question 3


```{r}
# Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

flights |>
  select(starts_with("dep"), starts_with("arr"))

flights |>
  select(ends_with("_time"), ends_with("_delay"),
         )
```

















